A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-28 paint-contour-tree.tex
minor section 1 changes
running time proof added

This instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
the painting procedure described above (which represents the time it takes to process this node).  
The path decomposition then turns out to be the heavy-light path decomposition 
described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
However, this decomposition is just the first step in the analysis, which 
requires further case analysis depending on the relative size of the length of each path 
and the weights of its vertices.

@@

Some basic facts about these trees. 
%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
vertices are splits, leaves are minima, and the global maximum is the root.
The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

@@

\smallskip
\begin{compactenum}
	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
	\item Add all candidates to queue.
	\item While candidate queue is non-empty: 
	\begin{compactenum}
		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
		consider edge in $\cS$. In either case, let edge by $(v,w)$.
		\item Insert $(v,w)$ in $\cA(\MM)$. 
		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
	\end{compactenum}
	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
\end{compactenum}
\end{minipage}}

@@

\begin{definition} The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
critical points that are \emph{not} non-dominant minima. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

@@

Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
Let us focus on an initial painting, where each colors has 1-1 correspondence with the maxima.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. A nice mental picture is given when $\MM$ is 2-dimensional.
In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
Moreover, each up-star of $v$ (one connecting to each mound) is fully colored. So this
indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
the parents of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds till $v$, so now that erstwhile join $v$
is a new maxima. Suppose purple paint was poured from $v$. In terms on $\jc(\MM)$, this
is equivalent to remove leaves $\ell_1$ and $\ell_2$, now making $v$ as a leaf.
Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.

@@

What happens when finally blue and red join at $v$? We merge the two colors, but now have
a blue heap and red queues of critical vertices. We also need to merge the priority queues
to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
critical points of one color. Merges can be done in logarithmic time, the same time
required for deletes.

In this discussion, we have ignored a somewhat annoying problem. Up-stars of critical vertices
may actually be touched by numerous colors, not just one as we assumed above. A simple solution
would be to insert vertices into heaps corresponding to all colors touching it. But there could
be super-constant numbers of copies of a vertex, and handling all these copies would lead
to an extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
one for each up-star. We are guaranteed that when $v$ needs to processed, its up-stars
automatically have one color, because of all the color merges that previously occured.

@@

\begin{claim} \label{clm:upstar} Consider ripe vertex $v$ and take the up-star connecting
to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
\end{claim}

\begin{proof} Let $c$ be the color of some edge in this up-star
By ripeness, $v$ is the highest in $T(\rep(c))$.
Denote the component of $\MM^+_v$ by $\PP$.
By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
\end{proof}

@@

In this section, we set up a framework to analyze the time taken to compute a merge tree.

\begin{definition}
 We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
 strictly increase along any root to leaf path.
\end{definition}

\begin{observation}
 The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
 to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
 factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
 manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
\end{observation}

Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
that technically the trees considered should have a leaf vertex hanging below the root of this 
full binary tree in order to represent the global minimum of the surface.  This vertex is 
(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.
We use $A(v)$ to denote the set of ancestors of $v$.
For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.

@@

\begin{lemma}
The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
% where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

\begin{proof} \Sesh{add}

>>>>

The instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
the painting procedure described above (which represents the time it takes to process this node).  
The path decomposition then turns out to be the heavy-light path decomposition 
described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
However, this decomposition is just the first step in the analysis, which 
requires further case analysis depending on the relative size of the length of each path 
and the weights of its vertices.

@@

Some basic facts about these trees. 
%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
vertices are splits, leaves are minima, and the global maximum is the root.
The key observation is that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.

@@

\smallskip
\begin{compactenum}
	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
	\item Add all candidates to queue.
	\item While candidate queue is non-empty: 
	\begin{compactenum}
		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
		\item Insert $(v,w)$ in $\cA(\MM)$. 
		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
	\end{compactenum}
	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
\end{compactenum}
\end{minipage}}

@@

\begin{definition} 
\label{def:criticalJoin}
The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
critical points other than the non-dominant minima. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

@@

Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. A nice mental picture is given when $\MM$ is 2-dimensional.
In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
Moreover, each up-star of $v$ (one connecting to each mound) is fully colored. So this
indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
the parents of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.

@@

What happens when finally blue and red join at $v$? We merge the two colors, but now have
blue and red queues of critical vertices. We also need to merge the priority queues
to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
critical points of one color. Merges can be done in logarithmic time, the same time
required for deletes.

In this discussion, we have ignored a somewhat annoying problem. Up-stars of critical vertices
may actually be touched by numerous colors, not just one as we assumed above. A simple solution
would be to insert vertices into heaps corresponding to all colors touching it. But there could
be super-constant numbers of copies of a vertex, and handling all these copies would lead
to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
one for each up-star. We are guaranteed that when $v$ needs to be processed, its up-stars
automatically have one color, because of all the color merges that previously occured.

@@

\begin{claim} \label{clm:upstar} Consider a ripe vertex $v$ and take the up-star connecting
to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
\end{claim}

\begin{proof} Let $c$ be the color of some edge in this up-star.
By ripeness, $v$ is the highest in $T(\rep(c))$.
Denote the component of $\MM^+_v$ by $\PP$.
By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
\end{proof}

@@

\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.
Also, the parent and child relationship is the opposite in this section as in earlier ones.}

In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
%
% \begin{definition}
%  We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
%  strictly increase along any root to leaf path.
% \end{definition}
% 
% \begin{observation}
%  The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
%  to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
%  factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
%  manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
% \end{observation}
%
As such any tree $T$ considered in this section will be a rooted binary tree\footnote{Note 
that technically the trees considered should have a leaf vertex hanging below the root of this 
in order to represent the global minimum of the complex.  This vertex is 
(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
length of the path between $v$ and $w$.
We use $A(v)$ to denote the set of ancestors of $v$.
For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.

@@

Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
according to the priority queue for each upstar we put a given vertex in.

\begin{lemma}
Let $\MM$ be a simplicial complex with $m$ edges.  For every vertex in $\jc(\MM)$, 
let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
The running time of our algorithm is $O(m\alpha(m) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
% where $T$ is the merge tree of the given input, 
%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
\end{lemma}

\begin{proof}
First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(m)$ time (note that as the input complex is connected, $N=O(m)$).
Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e. each edge is viewed at most twice).
Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.

Now consider the union-find operations performed by $\init$ and $\update$.  
Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
and so each union-find operation takes $O(m)$ time.
The $\update(K)$ procedure performs at most one find for each edge and no unions.  
$\init(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
operations is $O(m)$.

For the remaining operations, observe that every time $\update$ is called a vertex is pushed onto the stack and each
vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
Therefore the total running time due to $\update$ is linear (ignoring the find operations).

What remains is the time it takes process a vertex $v$ in $\init(\MM)$.  
In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
Therefore the only thing we must bound are the queue operations.
Let $v$ be a vertex in $\jc(\MM)$, and let $p_1$ and $p_2$ be its parents (the same argument holds if $v$ has only one parent).
At the time $v$ is handled, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
In particular, when $v$ is handled we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
these merged queues are the queues of the current colors of $p_1$ and $p_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
the only vertices of left in these queues are those in $H_{p_1}$ and $H_{p_2}$.

Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $p_1$ and $p_2$, and then the queues are 
are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{p_1}$, $H_{p_2}$, and $H_v$, respectively.  
As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
\end{proof}
